Investigating the correlation between a vote for Biden or Trump based on a persons education, gender, race and age.

### 2024-03-11 paper/paper.qmd
data section for paper

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Simulation
We first start with simulating the dataset, and recall that we are concerned with each voter's education level, gender, race and age. Our chosen sample size is 8000, and the categories for education and race have been named akin to the categories in the dataset. Thus, the header of the table obtained for the simulation is shown below:
```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false
# setting seed, prime number :P
set.seed(71)
library(dplyr)
# sample size 8000
num_obs <- 8000

@@

# preview the simulated data
cat("Table 1. Header for the political preferences dataset.")
head(us_political_preferences)
```

@@

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

Talk way more about it.

# Model

@@

>>>>

We are concerned with assessing whether we can predict what candidate a new survey respondent is going to vote for, given their age, education, gender and race. Accordingly, we shall be simulating a similar scenario for the 2020 US presidential election below.

# Simulation
We first start with simulating the dataset, and recall that we are concerned with each voter's education level, gender, race and age. Our chosen sample size is 8000, and the categories for education and race have been named akin to the categories in the dataset. Thus, the header of the table obtained for the simulation is shown below:
```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false
# setting seed, prime number :P
set.seed(71)
library(knitr)
library(tibble)
library(dplyr)
# sample size 8000
num_obs <- 8000

@@

# preview the simulated data
cat("Table 1. Header for the political preferences dataset.")
kable(head(us_political_preferences))
```

@@

We will be using the 2020 Cooperative Election Study (CES) (CITATION TODO) as our dataset, which can be previewed below:

```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false

library(dplyr)
library(tidyverse)

raw_data = read.csv("~/linear_model_investigation/data/ces2020_raw.parquet")
kable(head(raw_data))
```
To elaborate further on the dataset starting with the ```votereg``` column, values 1 and 2 signifies whether a person has voted or not, respectively. Naturally, for entries with a value of 2 under this column, the corresponding ```CC20_410``` value is NA, and in the other case the values 1 and 2 signify a vote for Biden or Trump respectively. The naming of this column corresponds to the question in the survey from which the data for it was collected, which asks "For whom did you vote for President of the United States?" (citation needed), with a total of 6 options, where Biden and Trump correspond to the first two.

Next, there are two values for gender: 1 and 2 corresponding to Male and Female. There are also a total of 8 categories under race corresponding to numbers 1 through 8, as well as 6 education categories ranging from "No HS" to "Post-grad".

Since we are interested with ages and not respondent birth years, we may process the dataset as such to reflect this:

```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false

raw_data$birthyr <- as.numeric(raw_data$birthyr)
current_year <- as.numeric(year(as.Date(Sys.time())))

# Subtract current year from birth year column to get age
raw_data <- raw_data %>% mutate(birthyr = current_year - birthyr)
# rename to age
names(raw_data)[names(raw_data) == "birthyr"] <- "age"
cat("Table 3. Header for 2020 US Election Survey data with the birthyr column converted to ages")
kable(head(raw_data))

# save data back into the csv.
write.csv(raw_data,"~/linear_model_investigation/data/ces2020.parquet")
```
We shall now filter for non NA values within the ```CC20_410``` column, as we are only concerned with respondents that cast a vote, which also means we filter for entries with a ```votereg``` value of 1. Additionally, we shall map each education level and race value to its corresponding value from the survey, and we finally obtain a table as follows:

```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false
# load in libraries.
library(tidyverse)
library(dplyr)
library(rstanarm)
library(modelsummary)

# read and convert values in each column to integer.
ces2020 <-
  read_csv(
    "~/linear_model_investigation/data/ces2020.parquet",
    col_types =
      cols(
        "votereg" = col_integer(),
        "CC20_410" = col_integer(),
        "gender" = col_integer(),
        "educ" = col_integer(),
        "race" = col_integer(),
        "age" = col_integer()
      )
  )

# formatting data, mapping race ids to races.
ces2020 <-
  ces2020 |>
  filter(votereg == 1,
         CC20_410 %in% c(1, 2)) |>
  mutate(
    voted_for = if_else(CC20_410 == 1, "Biden", "Trump"),
    voted_for = as_factor(voted_for),
    gender = if_else(gender == 1, "Male", "Female"),
    education = case_when(
      educ == 1 ~ "No HS",
      educ == 2 ~ "High school graduate",
      educ == 3 ~ "Some college",
      educ == 4 ~ "2-year",
      educ == 5 ~ "4-year",
      educ == 6 ~ "Post-grad"
    ),
    education = factor(
      education,
      levels = c(
        "No HS",
        "High school graduate",
        "Some college",
        "2-year",
        "4-year",
        "Post-grad"
      )
    ),
    race = case_when(
      race == 1 ~ "White",
      race == 2 ~ "Black",
      race == 3 ~ "Hispanic",
      race == 4 ~ "Asian",
      race == 5 ~ "Native American",
      race == 6 ~ "Middle Eastern",
      race == 7 ~ "Two or more races",
      race == 8 ~ "Other"
    ),
    race = factor(
      race,
      levels = c(
        "White",
        "Black",
        "Hispanic",
        "Asian",
        "Native American",
        "Middle Eastern",
        "Two or more races",
        "Other"
      )
    )
  ) |> select(voted_for, gender, education, race, age)
cat("Table 4. Header for Processed 2020 US Election Survey data.")
# preview data after mutation.
kable(head(ces2020))
```
# Model

@@

```{r}
#| include: true
#| warning: false
#| message: false
#| echo: false
cat("results")
political_preferences <-
  readRDS(file = "~/linear_model_investigation/data/political_preferences.rds")

# model summary
final_table = modelsummary(
  list(
    "Support Biden" = political_preferences
  ),
  statistic = "mad"
)
kable(final_table)
```
